Capturing the shape of our experiences and how we remember them

### 2018-08-20 paper/main.tex
discussion edits

\section{Discussion}
\label{sec:discussion}
Studying episodic memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall and item recognition experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words.

Here, we find that participants Historically, human memory reseachers have focussed on measuring the quantity of information recovered by an individual (e.g. the ``proportion correct'') under different experimental conditions. However, for real world experiences this is not a well-defined task. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, like the in the classroom or in a doctor's office.  Whereas instructors often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might miss the ``forrest for the trees''. True learning is about understanding the key concepts (i.e. understanding key themes in the learned content and how they relate) rather than about regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of a memory disorder, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the movie's topic trajectory? Or if participants were told that their narrations would be played back to other participants (cite Hasson work), would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing our" of the trajectories with each new retelling. This could be quantified using our topic trajectory approach.

%NOTE: could be useful to add a 'where this might break down' section

To conclude, let's revisit the question of what it means to remember something. We assert that episodic memories are not veridical snapshots of the past, and so we should stop treating (and modelling) them as such.  Rather, episodic memories are better described as echoes of a prior experience that are shaped by other past experiences as well as the current context.

>>>>

\section{Discussion}
\label{sec:discussion}
Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g. percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might miss the ``forrest for the trees''. We view ``true learning'' as understanding the key concepts (i.e. understanding key themes in the learned content and how they relate) rather than about regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of a memory disorder, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the movie's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal16}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g. one that describes the video in the same way as participant's recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g. participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to details (while maintaining flexibility).

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While its undeniable that these models have been immensely useful in advancing our understanding of the fundamental principles that govern human memory, they are severely limited in their ability to explain memory for naturalistic experiences. Real-life experiences are highly structured in time, and so our experiments and models of memory should account for this fact.
